---
title: "Riley 1946"
execute:
  freeze: auto
format:
  revealjs:
  #  theme: ["style.scss"]
    slide-number: c/t
    width: 1920
    height: 1080
highlight-style: monokai
---

```{r}
library(here)
library(deSolve)
library(patchwork)
library(mgcv)
library(tidyverse)
```

## Forcing

The forcing data needs to be interpolated between the observed data. Using `approx()` can lead to prblems, because i) linear interpolation is not suitable for all cases and ii) values requested for before the first or after the last lead to `NA` values (and stop the `ode()` function) or are assumed to be the same as the first or the last data point. This leads to "zagged"

Cubic splines or not?

```{r}

# load forcing data
dat_forc <- read_csv(here("data", "forcing_dat.csv"))

dat_forc %>%
  ggplot(aes(x = days, y = k)) +
  geom_line(col = "red") +
  geom_point(shape = 21, col = "red", fill = "white") +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x, k = 24, bs = 'cc'),
    se = F,
    alpha = .5
  ) +
  labs(x = "Time (days)", y = "Zooplankton biomass") +
  theme_light(base_size = 18)


```

## Forcing with code

Playing around with `approx()` for the forcing data, linear approximation does not seem a good choice for such circular data. Instead, GAMs with circular cubic splines will be used.

Define GAMs for each of the forcing variables

::::: columns
::: {.column width="50%"}

```{r}
#| echo: true

gam_list <- list() #<1>
vars <- c("I_0", "k", "L_N", "z_mix", "Temp", "zoo_bm")

for (v in vars) {
  gam_list[[v]] <- gam(
    formula = as.formula(paste(
      v,
      "~ s(days, k =",
      nrow(dat_forc),
      ")"
    )),
    data = dat_forc,
    method = "REML"
  )
}

```

1.  Take the `dat_penguins` data set, and then,

::: 
::: {.column width="50%"}
1.  Take the `dat_penguins` data set, and then,
:::
:::::

Set up parameters as in Riley 1946

## Paraneters

```{r}
parameters <- c(
  # parameters
  p = 2.5, # gC(gC)^-1 d^-1
  I_min = 0.0015, # g cal cm^-2 min^-1
  r = 0.069, # °C^-1 OR 0.69???
  R_0 = 0.0175, # day^-1
  g = 0.0075, # day^-1
  # forcing GAMs
  gam_list = gam_list
)
```

Define initial state and time axis

> Remark that initial condition is set to the value 15 days before the actual start of the model. Therefore, the start day is -15 days. This should work with GAMs, but not with `approx()`. Alternatively, the conditions at t = 0 days could be caluclated by running the model for 15 days.

```{r}
state <- c(Phyto_bm = 3.377) # initial condition
times <- seq(-15, 365, 1) # time axis
```

Define equations

```{r}
m_riley <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    I_0_t <- predict(gam_list.I_0, newdata = data.frame(days = time))
    k_t <- predict(gam_list.k, newdata = data.frame(days = time))
    L_N_t <- predict(gam_list.L_N, newdata = data.frame(days = time))
    z_mix_t <- predict(gam_list.z_mix, newdata = data.frame(days = time))
    Temp_t <- predict(gam_list.Temp, newdata = data.frame(days = time))
    zoo_bm_t <- predict(gam_list.zoo_bm, newdata = data.frame(days = time))

    # Euphotic depth
    z_euph <- log(I_0_t / I_min) / k_t

    # Light limitation
    L_V <- if (z_euph < z_mix_t) z_euph / z_mix_t else 1

    photos <- (p * I_0_t) /
      (k_t * z_euph) *
      (1 - exp(-k_t * z_euph)) *
      L_N_t *
      L_V

    resp <- R_0 * exp(r * Temp_t)
    graz <- g * zoo_bm_t

    dPhyto_bm <- (photos - resp - graz) * Phyto_bm

    list(
      dPhyto_bm,
      photos = photos,
      resp = resp,
      graz = graz,
      temp = Temp_t
    )
  })
}
```

Run the model

```{r}
solution <- ode(
  y = state,
  times = times,
  func = m_riley,
  parms = parameters
)
```

Plot Phytoplankton biomass

```{r}
plot_phytop <- solution %>%
  as_tibble() %>%
  filter(time > 0) %>%
  ggplot(aes(x = time)) +
  geom_area(aes(y = Phyto_bm), fill = "lightblue") +
  labs(x = "Days", y = expression("Phytoplankton biomass (gC " * m^-2 * ")")) +
  theme_light()

plot_phytop
```

Plot rates

> For some reason, `.1` is added to the variable names when using `ode()`. It seems to be related to passing the forcing variables as a list of GAMs and not as continuous vectors.

```{r}
solution %>%
  as_tibble() %>%
  filter(time > 0) %>%
  select(time, photos.1:graz.1) %>%
  # calculate rates
  mutate(
    Photosynthesis = photos.1,
    Respiration = Photosynthesis - resp.1,
    Grazing = Respiration - graz.1
  ) %>%
  # make a long df
  pivot_longer(
    cols = Photosynthesis:Grazing,
    names_to = "rate_type",
    values_to = "rate"
  ) %>%
  # adapt factor ordering for overlapping plotting
  mutate(
    rate_type = fct_relevel(
      rate_type,
      c("Photosynthesis", "Respiration", "Grazing")
    )
  ) %>%
  ggplot(aes(x = time, y = rate)) +
  geom_area(aes(fill = rate_type), position = "identity") +
  labs(
    x = "Days",
    y = expression("Carbon production (gC " * m^-2 ~ day^-1 * ")")
  ) +
  theme_light()
```

> When playing around using circular coords, the rates and biomass at 365 days is not the same as at 0 days. Probably due to starting conditions?

```{r}
data.frame(
  days = 1:365,
  I_0 = predict(parameters$gam_list.I_0, newdata = data.frame(days = 1:365))
) %>%
  ggplot(aes(x = days, y = I_0)) +
  geom_line() +
  geom_point(data = dat_forc, aes(y = I_0)) +
  theme_light()
```

```{r}
data.frame(
  days = 1:365,
  Temp = predict(parameters$gam_list.Temp, newdata = data.frame(days = 1:365))
) %>%
  ggplot(aes(x = days, y = Temp)) +
  geom_line() +
  geom_point(data = dat_forc, aes(y = Temp)) +
  theme_light()
```